Title TBD

### 2019-07-04 main/draft.tex
More paragraphs!

\begin{abstract}
Many couples marry shortly after woman gets pregnant, that is often referred as a shotgun marriage. I show that in American Community Survey data couples who experienced a shotgun marriage, i.e. had their first child before or at the year of their marriage, divorce more often in the future. This feature is robust in different subpopulations, being relatively more pronounced for wealthier people, although the share of shotgun marriages is lower among them. I argue the difference in divorce rates can be explained by endogenous selection into marriage: women who experienced unplanned pregnancy agree to enter marriages of lower quality, and in the data having kids before marriage is a plausible proxy for having lower-quality marriage. These patterns in selection can be rationalized by the observable structure of childcare costs, where kids require both money and time, and disproportional exposure of women to these costs. I estimate a structural model endogenizing marriage selection and use it to argue that marriage quality, as opposed to different composition of the groups, drives the major part of the observable difference in divorce probabilities. Qualitatively, this all suggests that the main reason for poor performance of shotgun marriages is large, unexpected and heterogeneous childcare costs. Quantitatively, the model suggests sizable effects of risks of unplanned pregnancy on observable divorce rates, especially in for young women.
\end{abstract}

@@

%[Paragraph 1: is shotgun marriage still a thing?]
Term ``shotgun marriage'' is traditionally used to describe situations when having a baby triggers marriage of a couple. This practice does not solely refer back to more conservative times: in the modern US data  around one-quarter of married women of age 21--40 had their first birth before or at the year of their first marriage\footnote{SIPP 2014}. And, although it seems that abortion and contraception access would make give couples more choice, in the data such couples still look like something forced them to marry each other. In particular, their marriages appear less stable: despite of being less than quarter of the population such couples account for more than one-third of divorces\footnote{ACS 2009--2017}.
%In this paper I argue that this pattern is rationalized by partners agreeing to enter marriages of lower quality as a response to unplanned pregnancy shock.

%[Paragraph 2: general importance? Child development]
The families who experienced shotgun marriages are often overlooked by the current literature focusing on child development. Growing up with a single mother or in a blended family (i.e. with step-parents) if conventionally considered as a predictor of poor early and adult-life outcomes of the children, and those who have two married biological parents are typically not considered to be at risk. However, not all marriages are created equal, and treating all marriages equally may understate inequality: growing up in unstable marriage can be a predictor of ending up with a single mother or in a different household, as well as factor of risk by its own.

@@

%[Paragraph 4: general importance? We need to understand trends]
On more aggregate level, this paper can broaden understanding of recent trends in household formation. First, around 40\% of new births in 2018 occurred by unmarried mothers. Second, increase in non-marital cohabitation is a documented trend, yet the median duration of cohabitation is 22 months\footnote{Source: Copen at al,  \cite{copen}}. Despite of this, unmarried unions with children are pretty rare: according to Payne, 2013\nocite{payne}, only about 3\% of children live in unions with unmarried cohabitating parents, as opposed to 21\% living with single mothers. Even mechanically this should suggest that large chunk of unmarried women transition into marriage shortly after they give births.

%[Paragraph 5: empirical findings --- main difference and subpopulations]
My main empirical contribution is describing association between having shotgun marriages and having a divorce. Although it is far from being causal, it is strong, robust to many ways of controlling for composition and is present among all subpopulations. In particular, [...]. The heterogeneity of this pattern can be informative of the channels and serves as a setup for the next model part of the paper.

%[Paragraph 6: empirical findings --- duration, step-children, lf and child development outcomes]
As additional findings, I establish few other results. [... ... ...]

%[Paragraph 7: model --- setup and why do we need a model]
To deal with obvious endogeneity and selection affecting the comparisons above I set up a structural model that is capable of understanding of anatomy of the result observed in the data. The model is based on growing branch of literature in dynamic household bargaining, and endogenizes savings, marriage and divorce decisions, fertility decisions and fertility shocks, together with the structure of childcare costs. The last element is crucial in generating heterogeneous response in terms of income or education: as time of wealthier parents is more expensive, unexpected childcare costs create relatively more distress for them. The key driving shock is that right before the moment people decide on their marriage they may unexpectedly get pregnant, and combination of it with differential costs of childcare generates desirable patterns in marriage selection. The model is estimated to match the key patterns of the data coming from American Community Survey, as well as supplementary sources, including ATUS and CEX for disciplining childcare costs part, and the structure I use fits the data extremely well.

%[Paragraph 8: model --- what I get from it, mechanics and anatomy]
The key conclusion of the structural part is that in estimated model couples who experienced an unplanned pregnancy shock agree to enter the marriages of the lower quality, and this is the major driver of the number that we observe in the data. he model delivers various decomposition, splitting observable cross-sectional difference in divorce rates onto effects of age, marriage length, income, wealth and marriage quality. As it turns out, the last factor is the most important contributor to the difference, and its contribution is [off the similar magnitude than the observable difference]. Decomposing the result further, the model predicts that the major part of the difference in divorce rates is driven by couples who would not have married if they did not have a baby (i.e. who complied with the shock), although there is some additional distress of having a baby on those who would have married anyway.

%[Paragraph 9: model --- counterfactuals and speculative things]
Additionally, the model delivers few nontrivial counterfactual predictions. First, in the model major part of divorces that happen early in life are driven by unexpected pregnancies, though divorces happening later are not. Second, general impact of pressure of becoming single mother on marriage decisions is substantial, and eliminating unexpected pregnancies happening before the marriage decision takes place lowers amount of people married by 25 by [...]. [Lastly, the model predicts decline in probabilities of unexpected pregnancies since 1980: the model matching fertility / marriage / divorce profiles from Census of 1980 predicts larger chances of unplanned pregnancies among all income and age groups.]

%[Paragraph 10: what I learned qualitatively]
As a general conclusion, the model teaches that conditional on enough observables, having kids before or at the year of marriage is a good proxy for an inferior marriage quality, and therefore the difference between two groups is not purely based on demography and economic outcomes. Therefore observing similar people with slight difference in fertility timing may be informative about the questions of how marital stability affects people's economic choices and future outcomes of them and their children.

%[Paragraph 11: what I learned about channels]
The causal lesson is that treat of being single mother is a major threat affecting how people make their marriage decisions, and large and unequally distributed costs of raising young children can be a driver in creation of poor quality marriages. This channel can be of major importance for studying welfare policies subsidizing childcare, young families and single mothers, as well as for enforcing alimony payments and child support in general. This also emphasizes the link between contraception and abortion access and marriages, although this impact is limited as even the wealthiest couples do experience sizable share of shotgun marriages, though this may be social rather than economic issue.

%[Paragraph 12: not a paragraph, just table of contents]
Section 2 of this paper discuss existing results and general trends that are already described by the literature. Section 3 discuss how I construct the dataset and timing variables, and Section 5 establishes main observations about differences in groups depending on their marriage timing, together with a number of additional considerations useful to understand the content of these differences. After that I explain theoretical model in Section 5 and the process of setting its parameters in Section 6. Section 7 describes estimation results and counterfactual simulations implied by these results. Section 8 provides more general discussion and concludes.

>>>>

\begin{abstract}
Many couples marry shortly after woman gets pregnant, that is often referred as a shotgun marriage. I show that in American Community Survey data couples who experienced a shotgun marriage, i.e. had their first child before or at the year of their marriage, divorce more often in the future. This feature is robust in different subpopulations, being relatively more pronounced for wealthier people, although the share of such marriages is lower among them. I argue this difference in divorce rates can be explained by endogenous selection into marriage: women who experienced unplanned pregnancy agree to enter marriages of lower quality. This implies that in the data having kids before marriage is a plausible proxy for having lower-quality marriage. Shortly, these patterns in selection can be rationalized by the structure of childcare costs, where kids require both money and time, and disproportional exposure of women to these costs. I estimate a structural model endogenizing marriage selection and use it to argue that differences in marriage quality, as opposed to different composition of the groups, drives the major part of the observable difference in divorce chances. Qualitatively, this suggests that the main reason for poor performance of shotgun marriages is large, unexpected and heterogeneous childcare costs. Quantitatively, the model suggests sizable effects of risks of unplanned pregnancy on observable divorce rates, especially for young women.
\end{abstract}

@@

%[Paragraph 1: is shotgun marriage still a thing?]
Term ``shotgun marriage'' is traditionally used to describe situations when having a baby triggers marriage of a couple. This practice does not solely refer back to more conservative times: in the modern US data around one-quarter of married women of age 21--40 had their first birth before or at the year of their first marriage\footnote{SIPP 2014}. And, although it seems that abortion and contraception access would provide people with more choice, in the data such couples still look like something pushed them to marry each other. In particular, their marriages appear less stable over time: despite of being less than quarter of the population shotgun marriage couples account for more than one-third of divorces\footnote{ACS 2009--2017}.
%In this paper I argue that this pattern is rationalized by partners agreeing to enter marriages of lower quality as a response to unplanned pregnancy shock.

%[Paragraph 2: general importance? Child development]
Understanding the sources of these difference is not just a household behavior question. The families who experienced shotgun marriages are often overlooked by the current literature focusing on child development. Growing up with a single mother or in a blended family (i.e. with step-parents) if conventionally considered\footnote{See Kearney, Levine 2017\nocite{kearney} for an extensive review of these issues} as a predictor of poor early and adult-life outcomes of the children, and those who have two married biological parents are typically not considered to be at risk. However, not all marriages are created equal, and treating all marriages equally may understate inequality: growing up in unstable marriage can be a predictor of ending up with a single mother or in a different household, as well as factor of risk by its own.

@@

%[Paragraph 4: general importance? We need to understand trends]
On more aggregate level, this paper can broaden the understanding of recent trends in household formation. First, around 40\% of new births in 2018 occurred by unmarried mothers. Second, increase in non-marital cohabitation is a documented trend, yet the median duration of cohabitation is 22 months\footnote{Source: Copen at al,  \cite{copen}}. Despite of this, unmarried unions with children are pretty rare: according to Payne, 2013\nocite{payne}, only about 3\% of children live in unions with unmarried cohabitating parents, as opposed to 21\% living with single mothers. Even mechanically this should suggest that large chunk of unmarried women transition into marriage shortly after they give births. Finally, this research can contribute to understanding differential behavior of marriage rates among racial and income groups, as the model identifies parameters of preferences and potential partners distribution, although it does not explicitly model cohabitation that seems to become a substitute for married unions.

%[Paragraph 5: empirical findings --- main difference and subpopulations]
I use American Community survey as the main dataset, as it has large sample sizes, uniform coverage and reports wide range of household and individual characteristics. One important drawback of it is that its marriage history is not reported perfectly, as it only has records of the year of the most recent marriage. I overcome this by conditioning on women who was married exactly once. I then use separate dataset of Survey of Income and Program Participation to re-establish the key findings without this restriction and show that this attrition does not introduce any important distortions besides changing magnitudes of what I refer to as divorce rates. Although SIPP provides superior coverage of the issues I study, the current work still mainly uses ACS: SIPP sample sizes do not allow to condition on multiple dimensions like child's age.

My main empirical contribution is describing association between having shotgun marriages and having a divorce. Namely, I partition population of women who were married and had kids on two groups:  those who gave their first birth at least at the next year of their marriage (marriage then kids, M$\to$K) and those who gave their first birth before or at the year of their marriage (kids then marriage, K$\to$M). I refer to the latter group as those who experienced a shotgun marriage. I additionally exclude from the comparisons women who had their birth more than five years before marriage, as they are increasingly likely to form a family with someone else than child's father.

%Although it is far from being causal, it is strong, robust to many ways of controlling for composition and is present among all subpopulations.

The empirical result is substantially larger share of divorce people in K$\to$M group. Although this contrast is far from being causal, it is strong, robust to many ways of controlling for composition and is present among all subpopulations. Namely, chances of being divorced at the moment of survey are around twice larger for those who experienced a shotgun marriage as opposed to for those who did not ($18\%$ against $10\%$). This ratio reaches three for those who have college degrees, and is only around 1.25 for those who only finished a high school. When controlled for many confounding factors that include age, income, education and geography, as well as marriage duration and child's age, the difference in shares vanishes around a half, but remains large and statistically significant.  %Similar heterogeneity holds if I use woman's income instead of education, although defining and measuring income can be subtle in case of women with young children.

%[Paragraph 6: empirical findings --- duration, step-children, lf and child development outcomes]
Investigating the data further leads to few additional insights. First, different timing of marriage and marriage duration at the moment of survey between the two groups does not seem to be an important contributor to the difference. Second, the divorces are concentrated around the time the child starts going to school. Third, for the cases when we can recover whether the child is step or own, this status does not affect the marriage duration, moreover, less than fifteen percent of the couples I observe seem to have step-children as I exclude those women who had their births long ago after their marriage. [SIPP multipartner fertility data confirm these findings?]
%As additional findings, I establish few other results. [... ... ...]

%[Paragraph 7: model --- setup and why do we need a model]
These comparisons, however, contain lots of selection issues and unobservable heterogeneity, and to deal with it I set up a structural model that is capable of understanding of anatomy of the result. The model is based on growing branch of literature in dynamic household bargaining, and endogenizes savings, marriage and divorce decisions, fertility decisions and fertility shocks, together with the structure of childcare costs. The last element is crucial in generating heterogeneous response in terms of income or education: as time of wealthier parents is more expensive, unexpected childcare costs create relatively more distress for them. The key driving shock is that right before the moment people decide on their marriage they may unexpectedly get pregnant, and combination of it with differential costs of childcare generates desirable patterns in marriage selection. The model is estimated to match the key patterns of the data coming from American Community Survey, as well as supplementary sources, including ATUS and CEX for disciplining childcare costs part, and the structure I use fits the data extremely well.

The main driving force in the model is an unplanned pregnancy shock. Each period single people meet potential partners with some probability, and if the partner is met, couple decides whether to marry each other based on a simple bargaining procedure. The potential matches are heterogeneous in partner's characteristics and marriage quality, the latter is modeled as idiosyncratic joint utility gain of couple from being together, that can be interpreted as ``love shock''.  Before the bargaining decision takes place, partners may or may not get an unplanned pregnancy shock, that proxies for unplanned children appearing in cohabitating unions as I do not model cohabitation explicitly. The shock interfers with bargaining: if the couple made a baby, woman risks to stay a single mother if the couple does not agree to marry. The conjecture is that in matches that experienced the shock women agree to enter the marriages of lower quality, the importance and even presence of this channel, however, depends on parameters of individual preferences and childcare costs that I estimate structurally.

%[Paragraph 8: model --- what I get from it, mechanics and anatomy]
The key conclusion of the estimation part is that couples who experienced an unplanned pregnancy shock indeed agree to enter the marriages of the lower quality, and this is the major driver of the number that we observe in the data. The model is capable of delivering various decompositions that split observable cross-sectional difference in divorce rates onto effects of age, marriage length, income, wealth and marriage quality. As it turns out, the last factor is the most important contributor to the difference, and its contribution is [of the similar magnitude than the observable difference]. Decomposing the result further, the model predicts that the major part of the difference in divorce rates is driven by couples who would not have married if they did not have a baby (i.e. who complied with the shock), although there is some additional distress of having a baby on those who would have married anyway.

%[Paragraph 9: model --- counterfactuals and speculative things]
Additionally, the estimated model delivers few nontrivial counterfactual predictions. First, major part of divorces that happen early in life are seen as driven by unexpected pregnancies, though divorces happening later are not. Second, general impact of pressure of becoming single mother on marriage decisions is substantial, and eliminating unexpected pregnancies happening before the marriage decision takes place lowers amount of people married by 25 by [...]. [Last, the model predicts decline in probabilities of unexpected pregnancies since 1980: the model matching fertility / marriage / divorce profiles from Census of 1980 predicts larger chances of unplanned pregnancies among all income and age groups.]

%[Paragraph 10: what I learned qualitatively]
As a general conclusion, the model teaches that conditional on enough observables, having kids before or at the year of marriage is a good proxy for an inferior marriage quality, and therefore the difference between two groups is not purely based on demography and economic outcomes. Therefore observing similar people with slight difference in fertility timing may be informative about the questions of how marital stability affects people's economic choices and future outcomes of them and their children, and, although this suggestion is model-based, many things in the actual data are consistent with it.

%[Paragraph 11: what I learned about channels]
The causal lesson is that being single mother is a major threat affecting how people make their marriage decisions, and large and unequally distributed costs of raising young children can be a driver in creation of poor quality marriages. This channel can be of major importance for studying welfare policies subsidizing childcare, young families and single mothers, as well as for enforcing alimony payments and child support in general. This also emphasizes the link between contraception and abortion access and marriages, although this impact is limited as even the wealthiest couples do experience sizable share of shotgun marriages, though this may be social rather than economic issue.

%[Paragraph 12: not a paragraph, just table of contents]
I elaborate on these points in the following order. Section 2 of this paper discuss existing results and general trends that are already described by the literature. Section 3 discuss how I construct the dataset and timing variables, and Section 5 establishes main observations about differences in groups depending on their marriage timing, together with a number of additional considerations useful to understand the content of these differences. After that I explain theoretical model in Section 5 and the process of setting its parameters in Section 6. Section 7 describes estimation results and counterfactual simulations implied by these results. Section 8 provides more general discussion and concludes.
